Cosmic Microwave Background Power Spectra from Few Bit Timestreams

### 2018-08-30 digitisation.tex
Editing and References.
Done more editing in line with what christian suggested. Began to include some references.

\begin{abstract}
Observations of the Cosmic Microwave Background (CMB) are of immense value to modern cosmology. However, future CMB experiments must confront challenges in mission planning, hardware and analysis that arise from the sheer size of the time-ordered-data (TOD) being recorded. These challenges are particularly significant for Antarctic and satellite experiments which depend on satellite links to transmit their data. We investigate using extreme digitisation to address these issues. Unlike lossless compression, extreme digitisation introduces additional noise into the data. We present optimal 1, 2 and 3 bit digitisation schemes and determine the degradation in temperature and polarisation power spectra caused by applying this process to the TOD. We find that 3 bit digitisation has a percent-level contribution to the map noise level. This is impressive considering that it would reduce the data volume by an order of magnitude. We argue that extreme digitisation is a promising strategy for upcoming experiments.
\end{abstract}

\keywords{ cosmic background radiation --- polarization --- data compression}
\section{Introduction}
\label{sec:intro}

Observations of the Cosmic Microwave Background (CMB) have played a key role in physics since 1964. Current and future CMB experiments will continue to deliver new insights by studying the temperature and polarisation information contained in the CMB. These will put tight constraints on cosmological models. The most prominent science goal is the discovery of the imprint of inflationary gravitational waves. Additionally, studies of CMB lensing and the Sunyaev-Zeldovich (SZ) effects will open up access to unprecedented insight. Moreover, through probing the relativistic number of species, the helium fraction and the neutrino mass sum, CMB experiments are a valuable counter-part to ground-based particle physics experiments.

The outstanding contribution of CMB science to modern physics has demanded a high standard of data analysis. The CMB community has developed a variety of compression and computational techniques to manage the increasing influx of data, while maximising the science output. These include the compression of time-ordered data (TOD) into maps, bandpower estimation and the pseudo $C_l$ method.

A growing hurdle for experiments at remote locations are the transmission limitations of satellite links. Space-based experiments have employed a combination of lossless and lossy compression techniques, including reduced bits in the TOD. Antarctica-based experiments that transmit a portion of their data via a satellite link have downsampled their data in the past to meet their telemetry requirements. They have yet to exploit few bit digitisation of the TOD. As we approach the next generation of ground-based experiments, Stage-4, and the launch of a new generation of space-based missions (liteBIRD, PIXIE, COrE+), we must treat the transmission bottleneck carefully. Without a review of the current compression techniques employed we are sure to lose information.

In this work we present the method of extreme digitisation, which compresses a rich digital input signal and compresses it into a few bits. We apply extreme digitisation to the TOD and detail its effect on temperature and polarisation power spectra. We find that an optimal 3-bit digitisation scheme adds as little as $< 2\%$ to the map noise level. While the digtisation schemes described here are primarily laid out for ground-based experiments, the results we present here should also be considered by future space-based missions, which inevitably must incorporate lossy compression.

This work is structured as follows. We detail the arising challenges in handling large TOD in \S\ref{subsec:problem}. We subsequently formulate extreme digitisation in \S\ref{subsec:extremedigitisation} and lay out the framework used to test its performance in \S\ref{subsec:method}. The details of the power spectrum estimation used are laid out in \S\ref{subsec:psestimation}. We continue by presenting the noise induced through the dicitisation process in \S\ref{subsec:additionalnoise} and summarise our findings in \S\ref{sec:conclusions}.

@@

The science goals of upcoming CMB experiments naturally lead to a large influx of data. To achieve the targeted sensitivity longer observations with more detectors are needed. In fact the number of detectors of ground-based experiments has been following a Moore's law like trend, doubling approximately every 2 years. This directly translates into an exponential growth in data volume.

The best observations sites for CMB measurements are at remote locations: in space and in the Antarctica. Experiments at these locations depend on satellite transmission. The next generation of ground-based experiments will aim to collect 2 million detector years of data. An experiment contributing to Stage-4 at the South Pole will face a data influx of $\sim \mathrm{O}(10)\mathrm{Tb/d}$. However, the current transmission allocation for SPT3G is at $150\mathrm{Gb/d}$, which will likely only see a moderate increase in coming years. The transmission bottleneck is currently overcome by recovering the full data on hard drives with some latency and by transmitting a downsampled version of the data. The downsampling process loses high frequency information. Going into Stage-4 we anticipate that compression rates for transmission must increase by an order of magnitude. Continued use of downsampling will narrow the information window decisively - prohibiting high multipole moment science to be carried out on the transmitted dataset. This also means that any potential faults or errors in the experiment that only become visible in high frequencies will go unnoticed for longer.

Future space-based missions aim to exceed the detector count of Planck by at least an order of magnitude. It is questionable whether their telemetry specifications will allow for transmission of the data with Planck-style compression. Methods of storing large amounts of data on upcoming satellites will likely be prohibited by financial decisions: the amount of storage space required becomes financially relevant at the scales targeted. Missions will likely be left to design their own compression algorithms which will incorporate a combination of lossless and lossy compression techniques.

@@

Beyond transmission challenges, mission planning is becoming exceedingly difficult. As noted by (S4 science book) a full simulation of TOD over the entire parameter space of detection scenarios for numerous set-ups is the desired way to decide on Stage-4 configurations. Space-based missions must aim to carry out a similar analysis to optimise their science output. Given the shear size of TOD expected, this is not possible. We must rely on different planning strategies or aim to reduce the size of the TOD in order to maximise the productivity of planning and development stages and guarantee scientific excellence.

@@

Operations on the TOD, such as noise-removal or map-making are a vital part of CMB data analysis. While we have not experienced the limitations of the accessible computational assets, the exponential growth of CMB data makes its analysis increasingly expensive.

@@

Extreme Digitisation would tackle the challenges mentioned above by reducing the size of the TOD by an order of magnitude. Together with already established lossless compression techniques (e.g. FLAC, run-length coding, Huffman coding, etc.) this will directly tackle transmission hurdles. While it needs to be investigated to what extent existing algorithms can be carried over, extreme digitisation has the potential of solving planning and analysis problems.

Other science areas have demonstrated that extreme digitisation is a valuable compression technique. Jenet and Anderson (cite) explored the application of such compression to radio pulsar timing measurements with success. Recently Clearwater et al. (in prep.) have investigated the advantages of using 1 and 2 bit data when searching for continuous gravitational waves using the Laser Interferometer Gravitational-wave Observatory (LIGO).

@@

% rephrase these, basically off Max1978

Digitisation is a lossy compression technique. However the induced noise depends on the number of bits used, the digitisation thresholds, and the output levels chosen. To minimise the noise induced through this process one must know the nature of the input signal. A theoretical framework to obtain these levels was laid out by Max in 1978. We review the key aspects of his work relevant for us below.

@@

This was carried out by Max. We incorporate his results by formulating the multi-level functions we use for our 1, 2 and 3 bit digitisation process. Given an input signal $s(t)$ a digitisation scheme using $N$ bits returns the output $\hat{s}_N(t)$. For 1 bit digitisation we apply the function

@@

Other digitisation schemes can be thought of, which place the digitisation thresholds and output levels in a different way. However, given the assumptions made the schemes derived above are optimal. Additionally, they are simple enough to be easily implemented computationally.

@@

To create the template maps we use the healpix framework and the wealth of support available for it. We generate a realisation of I, Q and U maps based on the results of Planck 2015. The key cosmological parameters are summarised in table \ref{tab:inputcosparams}.

We simulate observing a $\sim 600 \mathrm{deg^2}$ patch of the sky. To do so we perform a number of constant elevation scans (CES), equally spaced in declination (DEC). We repeat the observation strategy $100$ times with a slight offset in right ascension (RA) and DEC each time, such that all pixels within the patch are hit approximately uniformly. The speed at which we sweep across the survey area is adjusted to to produce a desired number of hits per pixel (hpp) in the output maps.

@@

This process is carried out 6 times, producing maps with different numbers of hpp. The simulation parameters are summarised in table \ref{tab:modelparams}. The calculation outlined above has considerable computational requirements if we want to reach up to $\sim 10^8$ hpp. To carry out this simulation we make use of parallelisation and the computing facilities provided by NERSC.

\begin{table}[tbh]
\begin{center}
\caption{\label{tab:inputcosparams} Input Cosmological Parameters}
\small
\begin{tabular}{l | c}
Parameter & Planck 2015\\
\hline\\[-1.5ex]
$100\theta_{MC}$ & $ 1.04086 \pm 0.00048$\\
$\Omega_b h^2$ & $  0.02222 \pm 0.00023$\\
$\Omega_c h^2$ & $ 0.1199 \pm 0.0022$\\
$H_0$ & $67.26 \pm 0.98$\\
$n_s$ & $0.9652 \pm 0.0062$\\
$\Omega_m$ & $0.316 \pm 0.014$\\
$\sigma_8$ & $0.830 \pm 0.015 $\\
$\tau$ & $0.078 \pm 0.019$\\
$10^9 A_s e^{-2\tau}$ & $1.881 \pm 0.014 $\\
\end{tabular}
\tablecomments{ 
Cosmological parameters used to create the template maps. Taken from Planck 2015: Cosmological Parameters.
} \normalsize
\end{center}
\end{table}

\begin{table}[tbh] % transpose and include hpp!
\begin{center}
\caption{\label{tab:modelparams} Assumed Survey Parameters}
\small
\begin{tabular}{l | c}
Parameter & Value\\
\hline \\[-1.5ex]
$\mathrm{NSIDE}$ &$4096$\\
$f_{\mathrm{readout}} [\mathrm{Hz}]$ &$200$\\
$f_{{\mathrm{sky}}}$ & $\sim 0.014$\\
$\sigma^{\mathrm{T}}_{\mathrm{det}} [\mathrm{\mu K \sqrt{s}}]$ &$500$\\
$\sigma^{\mathrm{Pol}}_{\mathrm{det}} [\mathrm{\mu K \sqrt{s}}]$ & $\sqrt{2}\times500$\\
\multirow{2}{*}{$\mathrm{hpp}$} & $\left(1, 10, 100 \right) \times 800,$\\
& $\left( 1, 10, 100 \right) \times 1024000$ \\
\end{tabular}
\tablecomments{ 
Parameters used in the simulated observation. The RA speed is adjusted to match the desired hpp.
} \normalsize
\end{center}
\end{table}

@@

We use PolSpice to compute the TT, EE and BB power spectra of the reconstructed maps I, Q, U maps. When doing so we apodise the observed skypatch using a cosine mask to minimise cut sky effects on the spectra. Please see table \ref{tab:polspiceparams} for an overview of the parameters used in this process.

\begin{table}[tbh]
\begin{center}
\caption{\label{tab:polspiceparams} PolSpice Parameters}
\small
\begin{tabular}{c c c c}
weights & $\sigma_{\mathrm{APOD}} [\mathrm{deg}]$ & apodistation type & polarisation \\
\hline
cosine mask & $\sqrt{600}/2$ & cosine & Yes \\

\end{tabular}
\tablecomments{ 
Parameters used when calling PolSpice to calculate the power spectra. Remaining parameters have been left at their default value.
} \normalsize
\end{center}
\end{table}

To normalise the the obtained power spectra we focus on the $\sim 10^8 \mathrm{hpp}$ maps. We normalise each power spectrum originating from few bit TOD against its corresponding control counterpart. We place a lower limit on the normalisation window in multipole space by considering the size of the observed patch with

\[ l = \frac{2}{\pi} \left( \frac{32400}{600} \right) \approx 35 \]

where we have rounded up to the next largest integer. We find the upper bound on the normalisation window by demanding that

\[ \frac{C_l^{\mathrm{S}}}{C_l^{\mathrm{N}}} \geq 10 \]

where $C_l^{\mathrm{S}}$ is the power spectrum of the input template maps and $C_l^{\mathrm{N}}$ the detector noise level.

we close the normalisation window the first instance the above condition is violated. The normalisation constants for each digitisation scheme and channel obtained through this way are then applied to all lower hpp simulations. A sample of the obtained power spectra are shown in figure \ref{fig:powspeczoom}.

\begin{figure}[htb]\centering
\includegraphics[width=0.5\textwidth,clip]{Plots/ttzoom.png}
\includegraphics[width=0.5\textwidth,clip]{Plots/eezoom.png}
\includegraphics[width=0.5\textwidth,clip]{Plots/bbzoom.png}
  \caption[Current ]{
   Reconstructed $\mathrm{TT}$, $\mathrm{EE}$, and $\mathrm{BB}$ power spectra. These originate from an observation with $102400000 \mathrm{hpp}$. We observe that digitisation appears to add some constant to the noise level. As one would expect $3 \mathrm{Bit}$ digitisation performs the best, followed by $2 \mathrm{Bit}$- and finally $1 \mathrm{Bit}$ digitisation.
\label{fig:powspeczoom}
}
\end{figure}

@@

The value of $l$ at which we can safely assume to be noise dominated and apply the above framework varies between simulated observations of different hpp and temperature and polarisation power spectra. The range considered involves any datapoints at multipole moments larger than the last point at which the detector noise is at least an order of magnitude larger than the template power spectrum, i.e.

@@

\begin{figure}[htb]\centering
\includegraphics[width=0.5\textwidth,clip]{Plots/ttratio.png}
\includegraphics[width=0.5\textwidth,clip]{Plots/eeratio.png}
\includegraphics[width=0.5\textwidth,clip]{Plots/bbratio.png}
  \caption[Current ]{
  Calculated ratio of $C_l^X/C_l^N$ of the rebinned $\mathrm{TT}$, $\mathrm{EE}$, and $\mathrm{BB}$ power spectra. The vertical black line indicates from which point onwards data is used to calculate the equivalent noise level. The above power spectra originate from a $80000\mathrm{hpp}$ map. Increasing the hpp in the map shifts the plateau to the right.
\label{fig:clxcln}
}
\end{figure}

The deduced additional noise for 1, 2 and 3 bit digitisation schemes are shown with respect to the hits per pixel in the maps in figure \ref{fig:extranoise}. We would like to point out three key results. Firstly, 3 Bit digitisation performs the best, followed by 2 Bit- and finally 1 Bit digitisation. This is what we expect, given that with each additional bit we retain more information. Secondly, for a fixed detector noise level the deterioration of the map quality scales in the same fashion as the number of hpp. We see that $\Delta \sigma / \sigma$ is independent of the hpp in the maps. Lastly, the added noise levels are astonishingly low. Keeping in mind that CMB detector sensitivity improves in steps of order of magnitude every few years, adding an extra percent-level noise term does not deteriorate the results appreciably. This is impressive, given that the use of an optimal 3 bit digitisation scheme will save approximately an order of magnitude in TOD volume.

@@

In this work we have motivated the investigation of extreme digitisation as a technique in combating arising data challenges in CMB data analysis. The reduction of the TOD by an order of magnitude directly addresses the issues in data transmission faced by remote location observations. Benefits in mission planning and data analysis are possible.

We have derived a set of optimal digitisation schemes and presented the level at which the induce noise into the temperate and polarisation power spectra. We find that an optimal 3 bit digitisation adds as little as $<2\%$ to the map noise level.

@@

%\bibliography{digitisation}

>>>>

\begin{abstract}

Observations of the Cosmic Microwave Background (CMB) are of significant value to modern cosmology and particle physics. Future CMB experiments face issues in hardware design, mission planning and analysis that arise from the vast size of time-ordered-data (TOD) being recorded. These challenges are particularly significant for Antarctic and satellite experiments which depend on satellite links to transmit their data. We explore the viability of reducing the TOD to few-bit numbers to address these issues. Unlike lossless compression, this technique introduces additional noise into the data. We present 1, 2 and 3 bit digitisation schemes and determine the degradation this compression causes in temperature and polarisation power spectra. We find that 3 bit digitisation has a percent-level contribution to the map noise level. We argue that such digitisation is a promising strategy for upcoming experiments.

%Observations of the Cosmic Microwave Background (CMB) are of immense value to modern cosmology. However, future CMB experiments must confront challenges in mission planning, hardware and analysis that arise from the sheer size of the time-ordered-data (TOD) being recorded. These challenges are particularly significant for Antarctic and satellite experiments which depend on satellite links to transmit their data. We investigate using extreme digitisation to address these issues. Unlike lossless compression, extreme digitisation introduces additional noise into the data. We present optimal 1, 2 and 3 bit digitisation schemes and determine the degradation in temperature and polarisation power spectra caused by applying this process to the TOD. We find that 3 bit digitisation has a percent-level contribution to the map noise level. This is impressive considering that it would reduce the data volume by an order of magnitude. We argue that extreme digitisation is a promising strategy for upcoming experiments.
\end{abstract}

\keywords{cosmic background radiation --- polarization --- data compression}
\section{Introduction}
\label{sec:intro}

Observations of the Cosmic Microwave Background (CMB) have played a key role in physics since its discovery \citep{penzias1965}. Current and future CMB experiments will continue to deliver new insights by studying the temperature and polarisation information contained in the CMB. These will put tight constraints on cosmological models. Upcoming science goals include the discovery of inflationary gravitational waves and thorough studies of CMB lensing and the Sunyaev-Zeldovich (SZ) effects. Moreover, CMB experiments will measure the relativistic number of species and the neutrino mass sum \citep{s4sciencebook}.

%The most prominent science goal is the discovery of the imprint of inflationary gravitational waves. Additionally, studies of CMB lensing and the Sunyaev-Zeldovich (SZ) effects will open up access to unprecedented insight. Moreover, through probing the relativistic number of species, the helium fraction and the neutrino mass sum, CMB experiments are a valuable counter-part to ground-based particle physics experiments.

High fidelity CMB science demands an excellent standard of data analysis. The CMB community has developed a variety of compression and computational techniques to manage the increasing influx of data, while maximising the science output \citep{tristam2007}. These include the compression of time-ordered data (TOD) into maps \citep{tegmark1997}, bandpower estimation \citep{tegmark1998} and the pseudo $C_l$ method \citep{brown2005}.

A growing hurdle for experiments at remote locations are the transmission limitations of satellite links. Space-based experiments have employed a combination of lossless and lossy compression techniques, including reduced bits in the TOD \citep{gaztanaga1998, maris2003}. Antarctica-based experiments that transmit a portion of their data via a satellite link have downsampled their data in the past to meet their telemetry requirements. They have yet to exploit few bit digitisation of the TOD. As we approach the next generation of ground-based experiments, Stage-4, and the launch of a new generation of space-based missions (liteBIRD, PIXIE, COrE+), we must treat the transmission bottleneck carefully. Without a review of the current compression techniques employed we are sure to lose information.

In this work we present the method of extreme digitisation, which compresses a rich digital input signal into a few bits. We apply extreme digitisation to the TOD and detail its effect on temperature and polarisation power spectra. We find that an optimal 3-bit digitisation scheme adds as little as $< 2\%$ to the map noise level. The digtisation schemes described here are laid out for ground-based experiments. Space-based missions must study carefully the effect of digitisation in their specific compression algorithms.% of t the results we present here should also be considered by future space-based missions, which inevitably must incorporate lossy compression.

This work is structured as follows. In \S\ref{sec:dig} we detail the challenges originating from handling large TOD, formulate the process of extreme digitisation and lay out the framework used to test its performance. Subsequently, we describe the power spectrum estimation employed and interpret the results obtained through this process in section \S\ref{sec:results}. We summarise our findings in \S\ref{sec:conclusions}.

%This work is structured as follows. We detail the arising challenges in handling large TOD in \S\ref{subsec:problem}. We subsequently formulate extreme digitisation in \S\ref{subsec:extremedigitisation} and lay out the framework used to test its performance in \S\ref{subsec:method}. The details of the power spectrum estimation used are laid out in \S\ref{subsec:psestimation}. We continue by presenting the noise induced through the dicitisation process in \S\ref{subsec:additionalnoise} and summarise our findings in \S\ref{sec:conclusions}.

@@

The science goals of upcoming CMB experiments naturally lead to a large influx of data To achieve the targeted sensitivity longer observations with more detectors are needed. In fact the number of detectors of ground-based experiments has been following a Moore's law like trend, doubling approximately every 2 years. This directly translates into an exponential growth in data volume \citep{s4sciencebook}..

Excellent observation conditions for CMB observations can not only be found in space, but also at various locations on earth \citep{li2017, kovac2007} among which is the South Pole. Space- and Antarctica-based experiments depend on satellite transmission. The next generation of ground-based instruments will aim to collect 2 million detector years of data. An experiment contributing to Stage-4 at the South Pole will face a data influx of $\sim \mathrm{O}(10)\mathrm{Tb/d}$. However, the current transmission allocation for SPT3G is at $150\mathrm{Gb/d}$, which will likely only see a moderate increase in coming years. The transmission bottleneck is currently overcome by recovering the full data on hard drives with some latency and by transmitting a downsampled version of the data. The downsampling process loses high frequency information. Going into Stage-4 we anticipate that compression rates for transmission must increase by an order of magnitude. Continued use of downsampling will narrow the information window decisively - prohibiting high multipole moment science to be carried out on the transmitted dataset. This also means that any potential faults or errors in the experiment that only become visible in said range will go unnoticed for longer.

The Planck mission has demonstrated the merits of carrying out CMB observations from space \citep{planck2018}. Upcoming missions aim to exceed the detector count of Planck by at least an order of magnitude \citep{litebird2014, pixie2011, core2018}. However, strategies to meet the telemetry specifications of each mission appear to be non-settled. It is not clear whether the data compression knowledge developed during the Planck mission will guarantee optimal performance for future satellites. Each mission will have to carefully construct a compression algorithm through a combination of lossless and lossy techniques.

%It is questionable whether their telemetry specifications will allow for transmission of the data with Planck-style compression. Methods of storing large amounts of data on upcoming satellites will likely be prohibited by financial decisions: the amount of storage space required becomes financially relevant at the scales targeted. Missions will likely be left to design their own compression algorithms which will incorporate a combination of lossless and lossy compression techniques.

@@

Beyond transmission challenges, mission planning is becoming exceedingly difficult. A full simulation of TOD over the entire parameter space of detection scenarios for numerous set-ups is the desired way to decide on Stage-4 configurations. Space-based missions must aim to carry out a similar analysis to optimise their science output. Given the shear size of TOD expected, this is not possible \citep{s4sciencebook}. We must rely on different planning strategies or aim to reduce the size of the TOD in order to maximise the productivity of planning and development stages and guarantee scientific excellence.

@@

Operations on the TOD, such as noise-removal or map-making are a vital part of CMB data analysis. While we have not experienced the limitations of the accessible computational assets, the exponential growth of CMB data makes its analysis increasingly expensive. The community depends on the excellent facilities provided by the National Energy Research Scientific Computing Center (NERSC).

@@

Extreme Digitisation would tackle the challenges mentioned above by reducing the size of the TOD by an order of magnitude. Together with already established lossless compression techniques (e.g. FLAC, run-length coding, Huffman coding, etc.) this will directly tackle transmission hurdles. Beyond that extreme digitisation has the potential of solving planning and analysis problems.

The Planck mission has demonstrated the advantages of using few bit compression of the TOD \citep{maris2003}. Other science areas have also shown the power of extreme digtisation. \citep{jenet1998} explored the application of such compression to radio pulsar timing measurements with success. Recently \citep{clearwater2018} have investigated the advantages of using 1 and 2 bit data when searching for continuous gravitational waves using the Laser Interferometer Gravitational-Wave Observatory (LIGO).

@@

Digitisation is a lossy compression technique. However, the induced noise depends on the number of bits used, the digitisation thresholds, and the output levels chosen. To minimise the noise induced through this process one must know the nature of the input signal. There is usually little concern around finding an optimal set of digitisation parameters for a given input signal. It is common to have access to large numbers of bits to store information, where changes in the digitisation scheme become insignificant to the distortion induced. It is in our interest however to consider extreme, i.e. few-bit, digitisation. To do so we review the key ideas laid out by \cite{max1960} below.

@@

We use the findings of Max to formulate the multi-level functions we use for our 1, 2 and 3 bit digitisation process. Given an input signal $s(t)$ a digitisation scheme using $N$ bits returns the output $\hat{s}_N(t)$. For 1 bit digitisation we apply the sign function

@@

The fact that the single bit digitisation scheme makes no references to the standard deviation of the signal is irrelevant. This affects the normalisation of the signal, which we can account for later. What is important is the spacing between digitisation thresholds and relative distance between the output levels chosen.

Completely different digitisation schemes can be thought of, which would for example reference the most drastic outlier in the dataset, or seek to place equal numbers of points into each digitisation level. However, given the assumptions made the schemes derived above are optimal. Additionally, they are simple enough to be easily implemented computationally.

@@

To create the template maps we use the healpix framework and the wealth of support available for it. We generate a realisation of I, Q and U maps with $\mathrm{NSIDE} = 4096$ based on the results of Planck 2015. The key cosmological parameters assumed are $\Omega_b \approx 0.049 $, $\Omega_c \approx 0.265$, $\Omega_m \approx 0.316$ and $h \approx 0.67$.

We simulate observing a $\sim 600 \mathrm{deg^2}$ patch of the sky. To do so we perform a number of constant elevation scans (CES), equally spaced in declination (DEC). We repeat the observation strategy $100$ times with a slight offset in right ascension (RA) and DEC each time, such that all pixels within the patch are hit approximately uniformly. The speed at which we sweep across the survey area is adjusted to to produce a desired number of hits per pixel (hits per pixel) in the output maps.

@@

This process is carried out 6 times. We produce maps with approximately $800$, $8,000$, $80,000$, $1,024,000$, $10,240,000$ and $102,400,000$ hits per pixel. We assume that the detector is read out $200$ times per second and has a noise level of $500 \mu K \sqrt{s}$ for temperature and $\sqrt{2} \times 500 \mu K \sqrt{s}$ for polarisation observations. The calculation outlined above has considerable computational requirements if we want to reach up to $\sim 10^8$ hits per pixel. To carry out this simulation we make use of parallelisation and the computing facilities provided by NERSC.

%\begin{table}[tbh]
%\begin{center}
%\caption{\label{tab:inputcosparams} Input Cosmological Parameters}
%\small
%\begin{tabular}{l | c}
%Parameter & Planck 2015\\
%\hline\\[-1.5ex]
%$100\theta_{MC}$ & $ 1.04086 \pm 0.00048$\\
%$\Omega_b h^2$ & $  0.02222 \pm 0.00023$\\
%$\Omega_c h^2$ & $ 0.1199 \pm 0.0022$\\
%$H_0$ & $67.26 \pm 0.98$\\
%$n_s$ & $0.9652 \pm 0.0062$\\
%$\Omega_m$ & $0.316 \pm 0.014$\\
%$\sigma_8$ & $0.830 \pm 0.015 $\\
%$\tau$ & $0.078 \pm 0.019$\\
%$10^9 A_s e^{-2\tau}$ & $1.881 \pm 0.014 $\\
%\end{tabular}
%\tablecomments{ 
%Cosmological parameters used to create the template maps. Taken from Planck 2015: Cosmological Parameters.
%} \normalsize
%\end{center}
%\end{table}

%\begin{table}[tbh] % transpose and include hits per pixel!
%\begin{center}
%\caption{\label{tab:modelparams} Assumed Survey Parameters}
%\small
%\begin{tabular}{l | c}
%Parameter & Value\\
%\hline \\[-1.5ex]
%$\mathrm{NSIDE}$ &$4096$\\
%$f_{\mathrm{readout}} [\mathrm{Hz}]$ &$200$\\
%$f_{{\mathrm{sky}}}$ & $\sim 0.014$\\
%$\sigma^{\mathrm{T}}_{\mathrm{det}} [\mathrm{\mu K \sqrt{s}}]$ &$500$\\
%$\sigma^{\mathrm{Pol}}_{\mathrm{det}} [\mathrm{\mu K \sqrt{s}}]$ & $\sqrt{2}\times500$\\
%\multirow{2}{*}{$\mathrm{hits per pixel}$} & $\left(1, 10, 100 \right) \times 800,$\\
%& $\left( 1, 10, 100 \right) \times 1024000$ \\
%\end{tabular}
%\tablecomments{ 
%Parameters used in the simulated observation. The RA speed is adjusted to match the desired hits per pixel.
%} \normalsize
%\end{center}
%\end{table}

@@

We use PolSpice to compute the TT, EE and BB power spectra of the reconstructed maps I, Q, U maps. When doing so we apodise the observed skypatch using a cosine mask with $\sigma_{\mathrm{APOD}} = \sqrt{600}/2 \mathrm{deg}$ to minimise cut sky effects on the spectra.

%\begin{table}[tbh]
%\begin{center}
%\caption{\label{tab:polspiceparams} PolSpice Parameters}
%\small
%\begin{tabular}{c c c c}
%weights & $\sigma_{\mathrm{APOD}} [\mathrm{deg}]$ & apodistation type & polarisation \\
%\hline
%cosine mask & $\sqrt{600}/2$ & cosine & Yes \\
%
%\end{tabular}
%\tablecomments{ 
%Parameters used when calling PolSpice to calculate the power spectra. Remaining parameters have been left at their default value.
%} \normalsize
%\end{center}
%\end{table}

Normalisation of the obtained power spectra is necessary. The digitisation schemes introduced in section \S\ref{sec:dig} are designed to minimise the distortion of the TOD, not to preserve power at the map level. Calibration to the CMB is common for many experiments (e.g. SPT).

For the normalisation procedure we focus on the $\sim 10^8 \mathrm{hits per pixel}$ maps. We normalise the $\mathrm{TT}$, $\mathrm{EE}$, and $\mathrm{BB}$ power spectra from few bit TOD against their control counterparts. We consider the normalisation window cut out by the conditions

\[ l > 35 \> ; \> \frac{C_l^{\mathrm{S}}}{C_l^{\mathrm{N}}} \geq 10 \]

where $C_l^{\mathrm{S}}$ is the power spectrum of the input template maps and $C_l^{\mathrm{N}}$ the detector noise level. The lower limit is placed in accordance to the patch-size surveyed, the upper limit ensures appropriate signal to noise in the normalisation window.

The normalisation constants for each digitisation scheme and channel are obtained through this way are then applied to all lower hits per pixel simulations. A sample of the obtained power spectra are shown in figure \ref{fig:powspeczoom}.

\begin{figure}[htb]\centering
\includegraphics[width=0.5\textwidth,clip]{Plots/ttzoom.png}
\includegraphics[width=0.5\textwidth,clip]{Plots/eezoom.png}
\includegraphics[width=0.5\textwidth,clip]{Plots/bbzoom.png}
  \caption[Current ]{
   Reconstructed $\mathrm{TT}$, $\mathrm{EE}$, and $\mathrm{BB}$ power spectra. These originate from an observation with $102400000 \mathrm{hits per pixel}$. We observe that digitisation appears to add some constant to the noise level. As one would expect $3 \mathrm{Bit}$ digitisation performs the best, followed by $2 \mathrm{Bit}$- and finally $1 \mathrm{Bit}$ digitisation.
\label{fig:powspeczoom}
}
\end{figure}

@@

The value of $l$ at which we can safely assume to be noise dominated and apply the above framework varies between simulated observations of different hits per pixel and temperature and polarisation power spectra. The range considered involves any datapoints at multipole moments larger than the last point at which the detector noise is at least an order of magnitude larger than the template power spectrum, i.e.

@@

\begin{figure}[htb]\centering
\includegraphics[width=0.5\textwidth,clip]{Plots/ttratio.png}
\includegraphics[width=0.5\textwidth,clip]{Plots/eeratio.png}
\includegraphics[width=0.5\textwidth,clip]{Plots/bbratio.png}
  \caption[Current ]{
  Calculated ratio of $C_l^X/C_l^N$ of the rebinned $\mathrm{TT}$, $\mathrm{EE}$, and $\mathrm{BB}$ power spectra. The vertical black line indicates from which point onwards data is used to calculate the equivalent noise level. The above power spectra originate from a $80000\mathrm{hits per pixel}$ map. Increasing the hits per pixel in the map shifts the plateau to the right.
\label{fig:clxcln}
}
\end{figure}

The deduced additional noise for 1, 2 and 3 bit digitisation schemes are shown with respect to the hits per pixel in the maps in figure \ref{fig:extranoise}. We would like to point out three key results. Firstly, 3 Bit digitisation performs the best, followed by 2 Bit- and finally 1 Bit digitisation. This is what we expect, given that with each additional bit we retain more information. Secondly, for a fixed detector noise level the deterioration of the map quality scales in the same fashion as the number of hits per pixel. We see that $\Delta \sigma / \sigma$ is independent of the hits per pixel in the maps. Lastly, the added noise levels are astonishingly low. Keeping in mind that CMB detector sensitivity improves in steps of order of magnitude every few years, adding an extra percent-level noise term does not deteriorate the results appreciably. This is impressive, given that the use of an optimal 3 bit digitisation scheme will save approximately an order of magnitude in TOD volume.

@@

In this work we have motivated the investigation of extreme digitisation as a technique in combating arising data challenges in CMB data analysis. The reduction of the TOD by an order of magnitude directly addresses the issues in data transmission faced by remote location observations. Benefits in mission planning, data analysis and hardware requirements are possible.

We have derived a set of optimal digitisation schemes and presented the level at which the induce noise into the temperate and polarisation power spectra. We find that an optimal 3 bit digitisation adds as little as $<2\%$ to the map noise level for temperature and polarisation observations. As mentioned above this addition is insignificant given that the sensitivity of CMB experiments follows a Moore's law like trend. The benefit of this compression technique is the reduction of the TOD volume by an order of magnitude.

@@

\bibliography{digitisation}
